Online {\IterComp}\\Guided by Low Overhead Profiling

### 2017-08-30 cgo18mscr.tex
minor updates
updates

\author{ \vspace{10em} }
\affiliation{%
  \institution{ }
}
\email{ }

@@

Each of these optimisations interacts with the code in complicated ways, depending on all other optimisations and the order they were applied to the code being optimised.
Understanding the interactions of optimisations is very important in determining a good solution to the phase-ordering problem~\cite{touati06,kulkarni12}.
Because of all those dependencies and interactions, although most compiler optimisations yield significant improvements in many programs, the potential for performance degradation in certain program patterns is known to compiler writers and many users~\cite{pan06,zhou12,kulkarni12}.
%Traditional compilers typically apply the same set of optimisation in a fixed order to all functions in a program, without regard the code being optimised.

Efficiently selecting the best combination of compiler optimisations for a given program or program section remains an open problem.
Compiler writers typically use a combination of experience and insight to construct the default sequences of prearranged optimisations found in compilers.
However, these default pre-arranged set of optimisations do not include all possible optimisations and they are always applied in the same pre-defined fixed order, without regard the code being optimised~\cite{pan06,cavazos07,zhou12,kulkarni12}.
%in hopes that a programmer will know which combination of optimisations will benefit their code~\cite{pan06,cavazos07,zhou12,kulkarni12}.

A well known compilation technique that addresses this challenge is {\itercomp}.
{\Itercomp} has the ability to adapt to new platforms, program and workload while still having a systematic and simple optimisation process.
It works by repeatedly evaluating a large number of compiler optimisations until the best combination is found for a particular program~\cite{fursin07,chen10}.
The main challenge concerning {\itercomp} is the need for efficiently exploring such a large optimisation space~\cite{fursin07,cavazos07,zhou12}.

Until recently, most of existing work  had been focusing on finding the best optimisation through repeated runs using a single input.
Although they demonstrate the potential of {\itercomp}, in real scenarios the user rarely executes the same input dataset multiple times~\cite{bodin98,kisuki99,stephenson03,kulkarni04,agakov06}.
Applying {\itercomp} in light of a single input may not result in good performance when executing the optimised code with different inputs.

Most of real world applications are complex enough so that a single input case does not capture the whole range of possible scenarios and program behaviour~\cite{haneda06,fursin07,chen10,chen12a}.
Because programs can exhibit behaviours that differ greatly depending on the input,
%This is something that is well known when writing test cases for correctness.
%We can extend the same argument when optimising the code for performance.
using a single input for {\itercomp} can produce poor performance when executed with different inputs.

Recent work have been studying the impact of using multiple input datasets
for performing {\itercomp}.
This previous work suggests that optimising based on a representative range of input datasets allows for selecting a robust compiler optimisation that produces good performance in real scenarios where the input is expected to change constantly~\cite{haneda06,fursin07,chen10,chen12a,chen12b,fang15,mpeis16}.
Their results show that a limited number of input datasets may be sufficient to obtain a well optimised program for a wider range of different inputs~\cite{haneda06,fursin07,chen10,chen12a}.

Finding such a robust combination of compiler optimisations by means of {\itercomp} across a large range of inputs may be very time consuming.
In order to speed up this process we intend to reduce the number of actual executions during the exploration of the optimisation space without much degradation of the final selected optimisation.

\begin{figure}[htb]
    \centering
    \includegraphics[width=\linewidth]{figs/infra-diagram}
    \caption{Overview of the execution engine for applying {\itercomp}.}
    \label{fig:infra-diagram}
\end{figure}

In this paper, our main goal is to enable {\itercomp} in \textit{online} scenarios with the restriction of executing distinct inputs only once, while targeting optimal performance across different inputs.
This online aspect is usually found in mobile and data centre platforms~\cite{chen12b,fang15,mpeis16}, where the goal is to optimise programs based on the workload of a particular user (or group of users) between executions.

Measuring just execution time, for example, is useful only if the amount of work is constant between executions with different inputs.
For this reason, we propose the use of a work-based metric in order to compare the performance of different optimisations across multiple executions of the program with distinct inputs.

Because of the restriction of not repeating inputs, we instrument the program for measuring the amount of work it performed during its execution.
Having a low overhead instrumentation is essential in this online scenario for two main reasons:
$(i)$ the user is directly affected by large overheads;
$(ii)$ a highly intrusive instrumentation can havea significant impact on the effect of the optimisations.

Our main contributions are the following:
\begin{itemize}
\item The use of a work-based performance metric in order to enable \textit{online} {\itercomp} by comparing different combination of compiler optimisations even when executed with distinct inputs.
\item We propose a relaxed instrumentation for low overhead profiling, with a controlled trade-off between accuracy and overhead.
\end{itemize}

\section{Work-based Metric} \label{sec:metric}

@@

%The hypothesis is that, given two optimisations $o_1$ and $o_2$, a program compiled with optimisation $o_2$ would \textit{consistently} perform better than when it is compiled with $o_1$, i.e., $o_1 \tilde{<}_p \, o_2$, if it performs more \textit{work} per unit of time when compiled with $o_2$ instead of $o_1$. 
By measuring the amount of \textit{work} done per unit of time, we reduce the impact of input-dependent aspects and focus instead on the efficiency of the optimised program.
For this metric, the main challenge is to precisely define what represents \textit{work}.
As discussed in Section~\ref{sec:related-metrics}, many work metrics have been presented in the literature, such as block frequency and others.
However, for comparing the performance benefits of different compiler optimisations on a program, although block frequency would be able to capture aspects of optimisations that simplify the control-flow graph (CFG), we argue that measuring work only at the basic block resolution would not be enough for capturing the effects of optimisations at the instruction level.
For this reason, we extend the idea of using basic block frequency to measure computational work by also considering the computational cost of each basic block.

%For that purpose, we intend to use profiling techniques to measure the input size and amount of computation performed on the execution path triggered by the input~\cite{ball94,ball96,zaparanuks12,coppa14}.
%In addition to being useful for speeding up {\itercomp} across a large number of different inputs, this metric based on comparing work-based efficiencies could
%potentially be used for applying {\itercomp} \textit{online}.
%However, the overhead for estimating the amount of \textit{work} must be low enough for the performance gains to be beneficial.
To this end, we model the computational work $\Delta W$ as a linear equation based on block frequency information and a cost model of the instruction set.
Formally,
\[
\Delta W = \varepsilon + \sum_{B} w(B)f(B)
\]
where $f(B)$ represents the frequency of basic block $B$ and $w(B)$ represents the computational work of executing $B$.
We define the work of a basic block $B$ as the sum of the cost of its instructions, i.e.,
\[
w(B) = \sum_{i} w_i N_B(i)
\]
where $w_i$ is the cost of instruction $i$ and $N_B(i)$ is the number of occurrences of instruction $i$ in basic block $B$.

In this simplified model, we consider that $w_i$ is constant across all programs and executions, varying only between target architectures.
On the other hand, $N_B(i)$ is a program-dependent static value which is known at compile-time and $f(B)$ is a dynamic value known only at run-time,
since $f(B)$ is both program and execution dependent as the execution frequency of a basic block can change when executing with different inputs.
%However, $N_B(i)$ is program dependent but constant across executions, while $f(B)$ is both program and execution dependent, since $f(B)$ can change when executing with different inputs.
%In other words, $N_B(i)$ is a static value known at compile-time and $f(B)$ is a dynamic value known only at run-time.

%If we define $x_i$ as
%\[
%x_i = \sum_{B\in P} N_B(i)f(B)
%\]
%we can write our definition of work $\Delta W$ as the following linear equation
%\[
%\Delta W = \varepsilon + \sum_{i\in I} w_i x_i
%\]

%\subsection{Linear regression}
%
%\begin{figure}[htb]
%    \centering
%    \includegraphics[width=\linewidth]{figs/cost-model.pdf}
%    \caption{Comparison between the naive and optimal instrumentation with no compiler optimisation.} %, i.e., compiled with \textbf{\texttt{-O0}}.}
%    \label{fig:cost-model}
%\end{figure}
%mse 0.000145373168444
%mae 0.00737261168981
%mae* 0.0033892281406
%mape 75.2924320971
%corr (0.99946441743545622, 0.0)

@@

We can illustrate this fact with a very small example as shown in Table~\ref{tab:ipc-example}.
This example shows two versions of a program, namely P1 and P2, and their respective measurements related to IPC.
Although version P1 has twice the IPC of P2, P1 is one cycle slower than P2.
As this example illustrates, the IPC metric can be misleading when compared different versions of the same program.
This problem is only intensified when both the optimisation sequence and the input changes.
In Chapter~\ref{chap:eval} we show empirical evidence for this argument.

\begin{table}[h]
\centering
\begin{tabular}{|c|c|c|c|}
\hline
                       & P1 & P2  \\
\hline
Number of Instructions & 5  & 2   \\
Number of Cycles       & 5  & 4   \\
IPC                    & 1  & 0.5 \\
\hline
\end{tabular}
\caption{Example that illustrates that a higher IPC does not necessarily translate to shorter execution time.}
\label{tab:ipc-example}
\end{table}

@@

LLVM is particularly suitable for iterative compilation as it makes possible to cache a pre-compiled, but still unoptimised, version of the input program in the bitcode format of the LLVM IR.  
This caching allows to speedup the time required for re-compilation as it is able to bypass the frontend phase.
If re-compilation time is critical, it would also be possible to keep only the hot portion of the code in the LLVM bitcode format, while the remaining portion of the code is already compiled to the final object code.
However, this is out of the scope of this thesis and we always re-compile the whole program.

@@

%In both phases the goal is to improve the performance of the execution phase, but from different perspectives.
%\textit{(i.)} In the compilation phase we focused on providing a low-overhead instrumentation for profiling the work metric. This phase is responsible to improve performance by lowering the overhead of the work profiling.
%\textit{(ii.)} In the evaluation phase we proposed the use of the WP metric in order to enable {\itercomp} in an online scenario.
%This phase is responsible to improve performance by being able to properly assess different optimisations such that the best optimisation can be selected.
%Chapter~\ref{chap:instr} describes our work profiling strategy and Section~\ref{sec:metric} describes the work-based metric.

%\textbf{Talk about the process of keeping the same optimization for a given input-window and how the optimizations as selected based on the evaluation of these "evidences". Perhaps we could suggest using the Theory of Evidences at this point?!}

@@

\begin{figure}[htb]
\centering {
  \includegraphics[scale=0.9]{figs/cfg-example.pdf}\\\vspace{1ex}
  \resizebox{0.45\textwidth}{!}{
  %\scalebox{0.8}{
     \begin{minipage}{0.5\textwidth}
     Instrumented value for each probe $P_i$:
     \begin{align*}
     w(P_0) &= w(B_0) + w(B_1) + w(B_4) + w(B_6) + w(B_7) + w(B_8) + w(B_9)\\
     w(P_1) &= w(B_2) - w(B_1) - w(B_4) - w(B_6) - w(B_7) - w(B_8)\\
     w(P_2) &= w(B_5) - w(B_6)\\
     w(P_3) &= w(B_3) - w(B_4) - w(B_6) - w(B_7)
     \end{align*}
     \end{minipage}
  }
}
  \caption{Example of a CFG with its minimum spanning tree in black and the
   basic blocks highlighted in red represent the instrumented basic blocks with
   the placement of the probes.}
  \label{fig:cfg-example}
\end{figure}

@@

\begin{figure}[h]
  \centering
  \includegraphics[scale=0.9]{figs/relax-instr-diagram.pdf}
  \caption{Overview of the work instrumentation algorithm, including the relaxation technique.}
  \label{fig:relax-instr-diagram}
\end{figure}

%\begin{figure*}[htb]
%    \centering
%    \includegraphics[width=\textwidth]{figs/overhead-O0.pdf}
%    \caption{Comparison between the naive and optimal instrumentation
%              with no compiler optimisation.}
%    \label{fig:overhead-O0}
%\end{figure*}

%Define an DAG of a CFG as a directed graph which may contain cycles if they have known constant trip counts.
Figure~\ref{fig:relax-instr-diagram} shows an overview of the relaxed instrumentation algorithm.
The highlighted step is introduced by the relaxed instrumentation on top of the previously defined optimal profiling.
The relaxed instrumentation performs a post-processing on the resulting instrumentation of the optimal algorithm.
This post-processing identifies probes that add little to the work metric, and removes their instrumentation.
In order to guarantee an upper bound for the dynamic error in the profiling measurement, the relaxation algorithm applies a constrained post-processing on a per DAG (directed acyclic graph) basis.
By constraining the relaxation within each DAG by a maximum allowed static error, it guarantees that the overall relaxation will also be constrained by the same bound.

The relaxation starts by extracting DAGs (directed acyclic graphs) from the CFG, as illustrated in Figure~\ref{fig:cfg-relax-example}.
First, the algorithm extracts all the subgraphs that represent a loop or the outer most region of the function.
Section~\ref{sec:natural-loops} explains how loops can be identified in the CFG.
Afterwards, these subgraphs are transformed into DAGs by ignoring the backedge and also by considering that any loop within the subgraph is never executed, i.e., only the headers of the inner loops are actually included into the DAG.
Figure~\ref{fig:cfg-relax-example} shows a CFG partitioned into two DAGs (consider only basic blocks and edges completely inside the yellow and green boundaries).
%
%LLVM has a Scalar Evolution Analysis which is used primarily to analyse expressions involving induction variables in loops~\cite{pop05}.
%An induction variable is a variable that is increased or decreased by a fixed amount on every iteration of a loop or is a linear function of another induction variable.
%
%This Scalar Evolution Analysis provides a way to compute \textit{small} constant trip counts (where \textit{small} is defined by a threshold).

\begin{figure}[h]
  \centering
  \includegraphics[scale=1]{figs/cfg-relax-example.pdf}
  \caption{Example of a CFG containing a loop and its decomposition into DAGs when applying the relaxation.
           The DAGs are the subgraphs within the dashed boundaries.}
  \label{fig:cfg-relax-example}
\end{figure}

\newpage
\begin{lstlisting}[caption={Optimal placement of probes for block frequency.}, label={lst:instrumentCFG}]
// Input: CFG	
relaxInstrumentation(G) {
  for loop in G:
     DAG = colapseInnerLoops(loop)
     relaxInstrumentedDAG(DAG)
  DAG = colapseInnerLoops(G)
  relaxInstrumentedDAG(DAG)
}
\end{lstlisting}

@@

We model the relaxation as a 0-1 Knapsack problem:
\begin{equation*}
\begin{aligned}
& \textrm{maximise }\quad \sum_{i=0}^{k} f(P_i)x_i \\
& \textrm{subject to }\quad \sum_{i=0}^{k} \varepsilon(P_i)x_i \leq M
\end{aligned}
\end{equation*}
\[
x_i\in\{0,1\}, i\in\{0,\ldots,k\}
\]
where $f(P_i)$ is the execution frequency of probe $P_i$, $x_i$ denotes the probes selected for removal, and $\varepsilon(P_i)$ is the percentage error of removing probe $P_i$ relative to the minimum work value possible to compute in the DAG, i.e., if $m$ is the minimum amount of work possible to be computed when executing the DAG, then $\varepsilon(P_i) = \frac{\omega(P_i)}{m}$.
Because the percentage error is computed based on the path with the minimum amount of work, $\varepsilon(P_i)$ represents the maximum error possible that would be incurred when removing probe $P_i$.
Furthermore, by constraining the percentage error of every DAG below a given threshold, we guarantee that the final error of the relaxation will always be bounded by the threshold, as demonstrated by Proposition~\ref{prop:relax-bound}.

\begin{prop}\label{prop:relax-bound}
Let $n_i$ be the number of times a given DAG $i$ is executed, $r_i$ be the total relaxation (amount of work removed) in DAG $i$, and $m_i$ be its minimum amount of work.
If $\frac{r_i}{m_i} \leq M$ for every $i$,
then the final error of the relaxation will always be bounded by the same threshold.
\end{prop}
\begin{proof}
We can model the overall error of the relaxation as:
\[
1 - \frac{n_1(m_1 - r_1) + n_2(m_2 - r_2) + \ldots + n_k(m_k - r_k) + c}{n_1m_1 + n_2m_2 + \ldots + n_km_k + c}
\]
That is,
%\begin{equation*}
\begin{gather*}
 1 - \frac{n_1m_1 + n_2m_2 + \ldots + n_km_k + c}{n_1m_1 + n_2m_2 + \ldots + n_km_k + c} + \frac{n_1r_1 + n_2r_2 + \ldots + n_kr_k}{n_1m_1 + n_2m_2 + \ldots + n_km_k + c} = \\
 \frac{n_1r_1 + n_2r_2 + \ldots + n_kr_k}{n_1m_1 + n_2m_2 + \ldots + n_km_k + c}
\end{gather*}
%\end{equation*}
If $\frac{r_j}{m_j}$ is the maximum ratio $\frac{r_i}{m_i}$ for every $i$, then
\begin{equation*}
\begin{aligned}
 \frac{n_1r_1 + n_2r_2 + \ldots + n_kr_k}{n_1m_1 + n_2m_2 + \ldots + n_km_k + c} &\leq\\
 \frac{n_1r_j + n_2r_j + \ldots + n_kr_j}{n_1m_j + n_2m_j + \ldots + n_km_j + c} &\leq\\
 \frac{n_1r_j + n_2r_j + \ldots + n_kr_j}{n_1m_j + n_2m_j + \ldots + n_km_j} & 
\end{aligned}
\end{equation*}
If $N = max\{n_i$ for every $i\}$, then
\begin{equation*}
\begin{aligned}
 \frac{n_1r_j + n_2r_j + \ldots + n_kr_j}{n_1m_j + n_2m_j + \ldots + n_km_j} &\leq\\
 \frac{Nr_j + Nr_j + \ldots + Nr_j}{Nm_j + Nm_j + \ldots + Nm_j} &=\\
 \frac{Nkr_j}{Nkm_j} = \frac{r_j}{m_j} &\leq M
\end{aligned}
\end{equation*}
\end{proof}

\begin{lstlisting}[caption={Optimal placement of probes for block frequency.}, label={lst:instrumentCFG}]
// Input: CFG	
relaxInstrumentedDAG(DAG){
   P = ProbesIn(DAG)
   m = minWork(DAG,P)
   K = createKnapsackModel(P,m)
   Bag = solveKnapsack(K)
   for B in (P-Bag):
     removeProbe(B)
}
\end{lstlisting}

@@

%In this section we discuss the three baseline implementations that will be used for assessing the quality of the proposed metric.

%\vspace{1ex}
%\noindent \textbf{{\Itercomp} over a single input}

%Most of the existing {\itercomp} studies find the best optimisation though repeated runs on the same input.
%Although this approach will usually lead to sub-optimal performance across large input datasets, it provides a good baseline when considering the complexity and resonably low compile-time regarding iterative optimising compilers.
%If $M$ is the total number of combinations of compiler optimisations, this approach requires $O(M)$ runs of the program being optimised.
%We can consider two main scenarios:
%\textit{($i$ - best case scenario)} after selecting the optimisation over each individual input, consider the one with best performance across the whole input dataset;
%\textit{($ii$ - expected scenario)} after selecting the optimisation over each individual input, consider the average case of their performance across the whole input dataset.

%\vspace{1ex}
%\noindent \textbf{{\Itercomp} across large input datasets}

>>>>

\author{ \vspace{1em} }
\affiliation{%
  \institution{ }
}
\email{ }

@@

Each of these optimisations interacts with the code in complicated ways, depending on all other optimisations and the order they are applied to the code being optimised.
However, understanding the interactions of optimisations is very important in determining a good solution to the phase-ordering problem~\cite{touati06,kulkarni12}.
Because of all those dependencies and interactions, although most compiler optimisations yield significant improvements in many programs, the potential for performance degradation in certain program patterns is known to compiler writers and many users~\cite{pan06,zhou12,kulkarni12}.
%Traditional compilers typically apply the same set of optimisation in a fixed order to all functions in a program, without regard the code being optimised.

Efficiently selecting the best optimisation sequence for a given program, or program section, remains an open problem.
Compiler writers typically use a combination of experience and insight to construct the default sequences of prearranged optimisations found in compilers.
However, these pre-arranged optimisation sequences do not include all possible optimisations and they are always applied in the same pre-defined order, without regard the code being optimised~\cite{pan06,cavazos07,zhou12,kulkarni12}.
%in hopes that a programmer will know which combination of optimisations will benefit their code~\cite{pan06,cavazos07,zhou12,kulkarni12}.

A well known compilation technique that addresses this challenge is {\itercomp}.
{\Itercomp} has the ability to adapt to new platforms, program and workload while still having a systematic and simple optimisation process.
It works by repeatedly evaluating a large number of optimisation sequences until the best combination is found for a particular program~\cite{fursin07,chen10}.
The main challenge concerning {\itercomp} is the need for efficiently exploring such a large optimisation space~\cite{fursin07,cavazos07,zhou12}.

Until recently, most of existing work  had been focusing on finding the best optimisation through repeated runs using a single input.
Although they demonstrate the potential of {\itercomp}, in real scenarios the user rarely executes a program with the same input multiple times~\cite{bodin98,kisuki99,stephenson03,kulkarni04,agakov06}.
Furthermore, most of real world applications are complex enough so that a single input case does not capture the whole range of possible scenarios and program behaviour~\cite{haneda06,fursin07,chen10,chen12a}.
Because programs can exhibit behaviours that differ greatly depending on the input,
%applying {\itercomp} in light of a single input may not result in a good performance when executing the optimised code with different inputs.
using a single input for {\itercomp} may not result in a good performance when executing the optimised code with different inputs.

%This is something that is well known when writing test cases for correctness.
%We can extend the same argument when optimising the code for performance.
%using a single input for {\itercomp} can produce poor performance when executed with different inputs.

Recent work have been studying the impact of using multiple inputs for
performing {\itercomp}.
%This previous work suggests that optimising based on a representative range of
%inputs allows for selecting a robust optimisation sequence that produces
%good performance in real scenarios where the input is expected to change
%constantly~\cite{haneda06,fursin07,chen10,chen12a,chen12b,fang15,mpeis16}.
Their results suggest that optimising based on a relatively small set of
representative inputs delivers good performance across a wide range of different
inputs~\cite{haneda06,fursin07,chen10,chen12a,chen12b,fang15,mpeis16}.
%Their results show that a limited number of inputs may be sufficient to obtain
%a well-optimised program for a wide range of different
%inputs~\cite{haneda06,fursin07,chen10,chen12a}.

In this work, our main goal is to enable {\itercomp} in \textit{online} scenarios.
For the purposes of this paper, we define the online scenario as having the restriction that programs execute multiple inputs and distinct inputs are executed only once.
This online aspect is usually found in mobile and data centre platforms~\citep{chen12b,fang15,mpeis16}, where the goal is to optimise programs based on the workload of a particular user (or group of users) between executions.
Online {\itercomp} must target optimal performance across different inputs.

Because of the restriction of having a single execution per input, it is not possible to measure speedup for comparing optimisations.
On the other hand, measuring just execution time, for example, is useful only if the amount of work is constant between executions with different inputs.
While previous work have suggested using \textit{instructions per cycle} (IPC) for performing {\itercomp} in \textit{online} scenarios, IPC have no correlation with speedup~\citep{fursin07}.

In order to address this challenge, we propose the use of a work-based metric to compare the performance of different optimisations across multiple executions of the program with distinct inputs.
We instrument the program for measuring the amount of work it performed during its execution.
Having a low overhead instrumentation is essential in this online scenario for two main reasons:
$(i)$ the user is directly affected by large overheads;
$(ii)$ a highly intrusive instrumentation can have a significant impact on the effect of the optimisations.
With the purpose of reducing profiling overhead, we propose two relaxation algorithms which provide a trade-off between measurement accuracy and overhead.
The first is a relaxation algorithm that operates on the level of regions of functions, while the second performs the relaxation considering the whole program at the same time.

Our experimental evaluation shows that performing online {\itercomp} guided by the work-based performance (WP) metric good results compared to the oracle, which is allowed to execute each input multiple times in order to use the actual speedup for guiding the {\itercomp}.
Online {\itercomp} guided by the WP metric is able to achieve an average of 7.5\% and a maximum of 33\% improvement over the standard {\flagstype -O3} optimisation.
Moreover, the experiments regarding the work profiling show that both relaxation algorithms are able to significantly reduce the profiling overhead while incurring a dynamic error of less than 5\% in the work measurement.
The whole program relaxation achieves an average of $2\times$ reduction in the overhead compared with the optimal profiling technique, while the more conservative relaxation that operates per region achieves an average improvement of 40\% over the optimal profiling.

%Our main contributions are the following:
%\begin{itemize}
%\item The use of a work-based performance metric in order to enable \textit{online} {\itercomp} by comparing different combination of compiler optimisations even when executed with distinct inputs.
%\item We propose a relaxed instrumentation for low overhead profiling, with a controlled trade-off between accuracy and overhead.
%\end{itemize}

To summarise, the main contributions of this paper are the following:
\begin{itemize}
\item We propose the use of a work-based performance (WP) metric for comparing different optimised versions of a program.
      In particular, we use the WP metric to guide {\itercomp} in an online scenario, where the program is expected to execute only once for distinct inputs.

\item Contrary to what previous work has suggested, we show that instructions per cycle (IPC) is not a good metric for online {\itercomp}.
\item We adapt the block frequency profiling in order to measure the WP metric.
\item A relaxation algorithm is proposed for lowering the overhead of the work profiling, with a controlled trade-off between accuracy and overhead.
\item A whole program relaxation is proposed in order to further reduce the overhead of the work profiling.
\end{itemize}

\section{Work-based Performance Metric} \label{sec:metric}

@@

By measuring the amount of \textit{work} done per unit of time, we reduce the impact of input-dependent aspects and focus instead on the efficiency of the optimised program.
For this metric, the main challenge is to precisely define what represents \textit{work}.
As discussed in Section~\ref{sec:related-metrics}, many work metrics have been presented in the literature, such as block frequency and others.
However, for comparing the performance benefits of different compiler optimisations on a program, although block frequency would be able to capture aspects of optimisations that simplify the control-flow graph (CFG), we argue that measuring work only at the basic block resolution would not be enough for capturing the effects of optimisations at the instruction level.
For this reason, we extend the idea of using basic block frequency to measure computational work by also considering the computational cost of each basic block.

To this end, we model the computational work $\Delta W$ as a linear equation based on block frequency information and a cost model of the instruction set.
Formally,
\[
\Delta W = \varepsilon + \sum_{B} w(B)f(B)
\]
where $f(B)$ represents the frequency of basic block $B$ and $w(B)$ represents the computational work of executing $B$.
We define the work of a basic block $B$ as the sum of the cost of its instructions, i.e.,
\[
w(B) = \sum_{i} w_i N_B(i)
\]
where $w_i$ is the cost of instruction $i$ and $N_B(i)$ is the number of occurrences of instruction $i$ in basic block $B$.

In this simplified model, we consider that $w_i$ is constant across all programs and executions, varying only between target architectures.
On the other hand, $N_B(i)$ is a program-dependent static value which is known at compile-time and $f(B)$ is a dynamic value known only at run-time,
since $f(B)$ is both program and execution dependent as the execution frequency of a basic block can change when executing with different inputs.

@@

% We can illustrate this fact with a very small example as shown in Table~\ref{tab:ipc-example}.
% This example shows two versions of a program, namely P1 and P2, and their respective measurements related to IPC.
% Although version P1 has twice the IPC of P2, P1 is one cycle slower than P2.
% As this example illustrates, the IPC metric can be misleading when compared different versions of the same program.
% This problem is only intensified when both the optimisation sequence and the input changes.
% In Chapter~\ref{chap:eval} we show empirical evidence for this argument.
%
% \begin{table}[h]
% \centering
% \begin{tabular}{|c|c|c|c|}
% \hline
%                        & P1 & P2  \\
% \hline
% Number of Instructions & 5  & 2   \\
% Number of Cycles       & 5  & 4   \\
% IPC                    & 1  & 0.5 \\
% \hline
% \end{tabular}
% \caption{Example that illustrates that a higher IPC does not necessarily translate to shorter execution time.}
% \label{tab:ipc-example}
% \end{table}

@@

LLVM is particularly suitable for iterative compilation as it makes possible to cache a pre-compiled, but still unoptimised, version of the input program in the bitcode format of the LLVM IR.
This caching allows to speedup the time required for re-compilation as it is able to bypass the frontend phase.
If re-compilation time is critical, it would also be possible to keep only the hot portion of the code in the LLVM bitcode format, while the remaining portion of the code is already compiled to the final object code.
However, this is out of the scope of this paper and we always re-compile the whole program.

@@

@@

\begin{figure}[htb]
\centering {
  \includegraphics[scale=0.85]{figs/cfg-example.pdf}\\\vspace{1ex}
  \resizebox{0.45\textwidth}{!}{
  %\scalebox{0.8}{
     \begin{minipage}{0.5\textwidth}
     Instrumented value for each probe $P_i$:
     \begin{align*}
     w(P_0) &= w(B_0) + w(B_1) + w(B_4) + w(B_6) + w(B_7) + w(B_8) + w(B_9)\\
     w(P_1) &= w(B_2) - w(B_1) - w(B_4) - w(B_6) - w(B_7) - w(B_8)\\
     w(P_2) &= w(B_5) - w(B_6)\\
     w(P_3) &= w(B_3) - w(B_4) - w(B_6) - w(B_7)
     \end{align*}
     \end{minipage}
  }
}
  \caption{Example of a CFG with its minimum spanning tree in black and the
   basic blocks highlighted in red represent the instrumented basic blocks with
   the placement of the probes.}
  \label{fig:cfg-example}
\end{figure}

@@

% \begin{figure}[h]
%   \centering
%   \includegraphics[scale=0.85]{figs/relax-instr-diagram.pdf}
%   \caption{Overview of the work instrumentation algorithm, including the relaxation technique.}
%   \label{fig:relax-instr-diagram}
% \end{figure}

%Figure~\ref{fig:relax-instr-diagram} shows an overview of the relaxed instrumentation algorithm.
%The highlighted step is introduced by the relaxed instrumentation on top of the previously defined optimal profiling.
The relaxed instrumentation performs a post-processing on the resulting instrumentation of the optimal algorithm.
This post-processing identifies probes that add little to the work metric, and removes their instrumentation.
In order to guarantee an upper bound for the dynamic error in the profiling measurement, the relaxation algorithm applies a constrained post-processing on a per DAG (directed acyclic graph) basis.
By constraining the relaxation within each DAG by a maximum allowed static error, it guarantees that the overall relaxation will also be constrained by the same bound.

The relaxation starts by extracting DAGs (directed acyclic graphs) from the CFG, as illustrated in Figure~\ref{fig:cfg-relax-example}.
First, the algorithm extracts all the subgraphs that represent a loop or the outer most region of the function.
Section~\ref{sec:natural-loops} explains how loops can be identified in the CFG.
Afterwards, these subgraphs are transformed into DAGs by ignoring the backedge and also by considering that any loop within the subgraph is never executed, i.e., only the headers of the inner loops are actually included into the DAG.
Figure~\ref{fig:cfg-relax-example} shows a CFG partitioned into two DAGs (consider only basic blocks and edges completely inside the yellow and green boundaries).

\begin{figure}[h]
  \centering
  \includegraphics[scale=0.85]{figs/cfg-relax-example.pdf}
  \caption{Example of a CFG containing a loop and its decomposition into DAGs when applying the relaxation.
           The DAGs are the subgraphs within the dashed boundaries.}
  \label{fig:cfg-relax-example}
\end{figure}

%\begin{lstlisting}[caption={Optimal placement of probes for block frequency.}, label={lst:instrumentCFG}]
%// Input: CFG
%relaxInstrumentation(G) {
%  for loop in G:
%     DAG = colapseInnerLoops(loop)
%     relaxInstrumentedDAG(DAG)
%  DAG = colapseInnerLoops(G)
%  relaxInstrumentedDAG(DAG)
%}
%\end{lstlisting}

@@

We model the relaxation as a 0-1 Knapsack problem:
\begin{equation*}
\begin{aligned}
& \textrm{maximise }\quad \sum_{i=0}^{k} f(P_i)x_i \\
& \textrm{subject to }\quad \sum_{i=0}^{k} \varepsilon(P_i)x_i \leq M
\end{aligned}
\end{equation*}
\[
x_i\in\{0,1\}, i\in\{0,\ldots,k\}
\]
where $f(P_i)$ is the execution frequency of probe $P_i$, $x_i$ denotes the probes selected for removal, and $\varepsilon(P_i)$ is the percentage error of removing probe $P_i$ relative to the minimum work value possible to compute in the DAG, i.e., if $m$ is the minimum amount of work possible to be computed when executing the DAG, then $\varepsilon(P_i) = \frac{\omega(P_i)}{m}$.
Because the percentage error is computed based on the path with the minimum amount of work, $\varepsilon(P_i)$ represents the maximum error possible that would be incurred when removing probe $P_i$.
 Furthermore, by constraining the percentage error of every DAG below a given threshold, we guarantee that the final error of the relaxation will always be bounded by the threshold.
% Furthermore, by constraining the percentage error of every DAG below a given threshold, we guarantee that the final error of the relaxation will always be bounded by the threshold, as demonstrated by Proposition~\ref{prop:relax-bound}.
%
% \begin{prop}\label{prop:relax-bound}
% Let $n_i$ be the number of times a given DAG $i$ is executed, $r_i$ be the total relaxation (amount of work removed) in DAG $i$, and $m_i$ be its minimum amount of work.
% If $\frac{r_i}{m_i} \leq M$ for every $i$,
% then the final error of the relaxation will always be bounded by the same threshold.
% \end{prop}
% \begin{proof}
% We can model the overall error of the relaxation as:
% \[
% 1 - \frac{n_1(m_1 - r_1) + n_2(m_2 - r_2) + \ldots + n_k(m_k - r_k) + c}{n_1m_1 + n_2m_2 + \ldots + n_km_k + c}
% \]
% That is,
% %\begin{equation*}
% \begin{gather*}
%  1 - \frac{n_1m_1 + n_2m_2 + \ldots + n_km_k + c}{n_1m_1 + n_2m_2 + \ldots + n_km_k + c} + \frac{n_1r_1 + n_2r_2 + \ldots + n_kr_k}{n_1m_1 + n_2m_2 + \ldots + n_km_k + c} = \\
%  \frac{n_1r_1 + n_2r_2 + \ldots + n_kr_k}{n_1m_1 + n_2m_2 + \ldots + n_km_k + c}
% \end{gather*}
% %\end{equation*}
% If $\frac{r_j}{m_j}$ is the maximum ratio $\frac{r_i}{m_i}$ for every $i$, then
% \begin{equation*}
% \begin{aligned}
%  \frac{n_1r_1 + n_2r_2 + \ldots + n_kr_k}{n_1m_1 + n_2m_2 + \ldots + n_km_k + c} &\leq\\
%  \frac{n_1r_j + n_2r_j + \ldots + n_kr_j}{n_1m_j + n_2m_j + \ldots + n_km_j + c} &\leq\\
%  \frac{n_1r_j + n_2r_j + \ldots + n_kr_j}{n_1m_j + n_2m_j + \ldots + n_km_j} &
% \end{aligned}
% \end{equation*}
% If $N = max\{n_i$ for every $i\}$, then
% \begin{equation*}
% \begin{aligned}
%  \frac{n_1r_j + n_2r_j + \ldots + n_kr_j}{n_1m_j + n_2m_j + \ldots + n_km_j} &\leq\\
%  \frac{Nr_j + Nr_j + \ldots + Nr_j}{Nm_j + Nm_j + \ldots + Nm_j} &=\\
%  \frac{Nkr_j}{Nkm_j} = \frac{r_j}{m_j} &\leq M
% \end{aligned}
% \end{equation*}
% \end{proof}

%\begin{lstlisting}[caption={Optimal placement of probes for block frequency.}, label={lst:instrumentCFG}]
%// Input: CFG
%relaxInstrumentedDAG(DAG){
%   P = ProbesIn(DAG)
%   m = minWork(DAG,P)
%   K = createKnapsackModel(P,m)
%   Bag = solveKnapsack(K)
%   for B in (P-Bag):
%     removeProbe(B)
%}
%\end{lstlisting}

@@
